---
title: "Space-Time Prediction of Bike Share Demand: Philadelphia Indego"
author: "Hope Levin & Annalise Abraham"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

# Setup

## Load Libraries

```{r load_libraries}
# Core tidyverse
library(tidyverse)
library(lubridate)
library(dplyr)

# Spatial data
library(sf)
library(tigris)

# Census data
library(tidycensus)

# Weather data
library(riem)  # For Philadelphia weather from ASOS stations

# Visualization
library(viridis)
library(gridExtra)
library(knitr)
library(kableExtra)

# here!
library(here)
# Get rid of scientific notation.
options(scipen = 999)
```

## Define Themes

```{r themes}
plotTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 11, face = "bold"),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "#D0D0D0", size = 0.2),
  panel.grid.minor = element_blank(),
  axis.ticks = element_blank(),
  legend.position = "right"
)

mapTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8),
  axis.line = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  panel.grid.major = element_line(colour = 'transparent'),
  panel.grid.minor = element_blank(),
  legend.position = "right",
  plot.margin = margin(1, 1, 1, 1, 'cm'),
  legend.key.height = unit(1, "cm"),
  legend.key.width = unit(0.2, "cm")
)

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

## Set Census API Key

```{r census_key, eval=FALSE}

census_api_key("5a82e243438bea307ae1c04f150d539c4db5fa47", overwrite = TRUE, install = TRUE)

```

```{r census_key_hidden, include=FALSE}
# Hidden key for rendering
census_api_key("5a82e243438bea307ae1c04f150d539c4db5fa47")
```

------------------------------------------------------------------------

# PART 1.1: Data Import & Preparation

## Load Indego Trip Data (Q2 2024)

```{r load_indego}
# Read Q2 2024 data
indego <- read_csv("indego-trips-2024-q2.csv")
```

## Create Time Bins

```{r create_time_bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

```

We choose to examine Quarter 2 (April - June) in 2024, as it reflects the start of nice biking weather for 2024. Additionally, it was the farthest from Quarter 1 in 2025, which we believed could be interesting to examine.

------------------------------------------------------------------------

# PART 1.2: Exploratory Analysis

## Trips Over Time

```{r trips_over_time}
# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2024",
    subtitle = "Spring Demand Patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego Bike Share"
  ) +
  plotTheme
```

Daily trip volumes rose from early April to mid-June. Mid-June to July marked a slight downfall, which we believe to be a result of riders opting other transit methods due to high temperatures. Regardless, the overall trend can be described as a steady increase. Compared to Quarter 1 in 2025, the number of daily trips is much higher. We believe this to be a product of people opting to bike in due to warmer/less harsh climate patterns. These persons could be commuters, or those looking for a fun and inexpensive outdoor activity.

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear Commute Patterns on Weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

Hourly weekday ridership patterns indicate a rush hour peak time-based pattern. From the graph, we observed that peak weekday ride times are around 7am and 5pm. Hourly weekend ridership patterns are much different, with a gradual rise to about mid-day, with a gradual decline.

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The top station reported 6,115 trips. The next highest station reported 5,231, a 884 decline. Concluding, the 20th most station reported 3,404. 13/20 of the top stations fell within the 3,000 - 3,999 range.

------------------------------------------------------------------------

# Get Philadelphia Spatial Context

## Load Philadelphia Census Data

```{r load_census}
# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  # WGS84 for lat/lon matching
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for Understanding Bike Share Demand Patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

This map reveals that most Indego stations are concentrated in census tracts with a high median income, like Center City and Old City. However, many tracts are located in lower income tracts, like University City.

## Join Census Data to Stations

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Stations in Non-Residential Census Tracts
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% dplyr::select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      dplyr::select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% dplyr::select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego Stations Shown (RED = No Census Data Match)",
    caption = "Red Xs indicate stations that did not join to census tracts"
  ) +
  mapTheme
```

Majority of the Indego stations joined to their respective census tract. However, some stations in what appears to be University City and the Sports Complex area (?) did not join. Because these are areas that could be categorized as institutional or recreational, we believe that demand cannot be based on demographics. To elaborate, few people live in the Sports Complex area- but many travel there.

# Dealing with Missing Data

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      dplyr::select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )
```

# Get Weather Data

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2024-04-01",
  date_end = "2024-06-30"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  dplyr::select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct()

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")
```

## Visualize Weather Patterns

```{r visualize_weather}
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2024",
    subtitle = "Spring to Early Summer Transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

Overall, temperatures rose in steady ebbs and flows from April into July. By July, temperatures were consistently over 80F, a big jump from the sub-50F temperatures in early April.

------------------------------------------------------------------------

# Create Space-Time Panel

## Aggregate Trips to Station-Hour Level

```{r aggregate_trips}
# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))
```

## Create Complete Panel Structure

```{r complete_panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand.grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel, by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

# Verify complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")
```

------------------------------------------------------------------------

# Create Temporal Lag Variables

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past Demand Predicts Duture Demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

Based on this visualization, we determined that hours with activity are typically followed by another hour of activity. This is called short-term persistence, or smooth demand changes.

------------------------------------------------------------------------

# Temporal Train/Test Split

```{r temporal_split}

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)

# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 23)

test <- study_panel_complete %>%
  filter(week >= 23)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")
```

------------------------------------------------------------------------

# Build Predictive Models

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

------------------------------------------------------------------------

# PART 1.3: Model Evaluation

## Calculate Predictions and MAE

```{r calculate_mae}
# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Model 1 - Time + Weather This model can best be described as a basic baseline model that demonstrates how trip counts can be impacted by weather time. The R\^2 was 0.11 meaning that Model 1 accounts for 11% of variations in trip counts. With an MAE score of .87, the model is not appropriate for predictions.

Model 2 - Time + Weather + Temporal Lags This model added lagged trip counts, which could be used to determine if past demand predicts future demand. We found that use is correlated with time, with rush hours at 7am and 5pm as proof. The R\^2 was 0.34 meaning Model 2 accounts for 34% of variations in trip counts. However, this lag allowed for better predictions, as the MAE fell to 0.71.

Model 3 - Time + Weather + Temporal Lags + Demographics This model added socioeconomic data like median income to see how demographic data impacts bike use patterns. The R\^2 was 0.23 meaning Model 3 accounts for 23% of variations in trip counts. MAE was 0.97 - aggressively high for making predictions.

Model 4 - Time + Weather + Temporal Lags + Demographics + Station Fixed Effects This model added station fixed effects, which controlled for unobserved differences between stations. The R\^2 was 0.26 meaning Model 4 accounts for 26% of variations in trip counts. The MAE was 0.95. Both R\^2 and MAE were slight improvements from Model 3, but overall still a poor performing model.

Model 5 - Time + Weather + Temporal Lags + Demographics + Station Fixed Effects + Rush Hour Interactions This model added rush hour interactions. From this, we learned that peaks differ on weekends and weekdays, and trip count varies based on month due to temperature/seasonal changes. The R\^2 and MAE were the same as Model 4.

Concluding, Model 2 is the best of the 5, but requires work to improve predictions.

Looking to Quarter 1 2025, we found that demand is higher in Quarter 2 2024 due to the introduction of warmer weather. However, given transition between seasons, Quarter 2 suffers from highs and lows in use. Both Quarter 1 2025 and Quarter 2 2024 had similar MAE scores. Moving forward, we now know that ridership is higly dependent on temperature, which we cannot predict, especially due to irregular weather patterns from climate change.

------------------------------------------------------------------------

# PART 2.1: Space-Time Error Analysis

## Observed vs. Predicted

```{r obs_vs_pred}
test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 Performance by Time Period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

## Spatial Error Patterns

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat.x, start_lon.y) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat.x), !is.na(start_lon.y))

## Create Two Maps Side-by-Side with Proper Legends

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat.x, start_lon.y) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat.x), !is.na(start_lon.y))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
grid.arrange(
  p1, p2,
  ncol = 2
  )
p1
```

Prediction errors are highest in Center City, which we know to be the area with highest demand, an attribute that stems from it having the most stations and highest ridership (and perhaps the most protective bike lanes). Because Center City is home to many landmarks and office buildings, much of its data is impacted by non-residents/tourists and commuters. Other neighborhoods have a lower MAE because they have stable use, a factor we believe stems from majority use by residents. Continuing, a high MAE and high demand score indicate that the model's poor performance comes from high variability

## Temporal Error Patterns

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Regardless of day type, the model struggles most during the PM Rush with an MAE score above 1 for both weekdays and weekends- awful predictive performance! Overnight weekday and weekend MAE scores were the lowest, with figures just above 0.25, suggesting strong predictive performance.

## Errors and Demographics

```{r errors_demographics}
# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% dplyr::select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

grid.arrange(p1, p2, p3, ncol = 2)
```

Errors arose when incorporating median income and percent of population white. For example, a median income of \$150,000 meant an MAE score of \~1.25 and 75% white population meant an MAE score of \~1.15. When you went back to 0 for both, the MAE scores were \~0.75. However, the model performed much better as percent of population taking transit rose. In short, the model performs poorly in richer and/or whiter areas, and performs better in transit heavy areas. Concluding, we now know the model does not suffer from underperformance due to systematic biases.

------------------------------------------------------------------------

# PART 3.1: Feature Engineering & Model Improvement

## Added Feature #1 (HL: I SUGGEST USING PERFECT BIKING WEATHER AND SCHOOL CALENDAR (APRIL-MAY) - TEXT ME IF YOU HAVE QUESTIONS)

```{r}

```

## Added Feature #2

```{r}

```

------------------------------------------------------------------------

# PART 4.1: Critical Reflection

------------------------------------------------------------------------

STEPS WE HAVE LEFT: OH WE ALSO HAVE TO MAKE THIS AN RMD FILE MY BAD

Based on your error analysis, add 2-3 NEW features to improve the model:

**Potential features to consider:**

*Temporal features:*

-   Holiday indicators (Memorial Day, July 4th, Labor Day)
-   School calendar (Penn, Drexel, Temple in session?)
-   Special events (concerts, sports games, conventions)
-   Day of month (payday effects?)

*Weather features:*

-   Feels-like temperature (wind chill/heat index)
-   "Perfect biking weather" indicator (60-75°F, no rain)
-   Precipitation forecast (not just current)
-   Weekend + nice weather interaction

*Spatial features:*

-   Distance to Center City
-   Distance to nearest university
-   Distance to nearest park
-   Points of interest nearby (restaurants, offices, bars)
-   Station capacity
-   Bike lane connectivity

*Trip history features:*

-   Rolling 7-day average demand
-   Same hour last week
-   Station "type" clustering (residential, commercial, tourist)

**Implementation:**

-   Add your features to the best model
-   Compare MAE before and after
-   Explain *why* you chose these features
-   Did they improve predictions? Where?

**Try a poisson model for count data** - Does this improve model fit?

------------------------------------------------------------------------

## Part 4: Critical Reflection

Write 1-2 paragraphs addressing:

1.  **Operational implications:**
    -   Is your final MAE "good enough" for Indego to use?
    -   When do prediction errors cause problems for rebalancing?
    -   Would you recommend deploying this system? Under what conditions?
2.  **Equity considerations:**
    -   Do prediction errors disproportionately affect certain neighborhoods?
    -   Could this system worsen existing disparities in bike access?
    -   What safeguards would you recommend?
3.  **Model limitations:**
    -   What patterns is your model missing?
    -   What assumptions might not hold in real deployment?
    -   How would you improve this with more time/data?
